---
title: "Exploring Your Data"
author: "zoontutorials team"
date: "4 April 2017"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Exploring Your Data}
  %\VignetteEncoding{UTF-8}
---

## Data Exploration
Data exploration is a key first step before you begin to think about what kind of model module you will use. 

```{r message = F, warning = F}
library(zoon)
library(lattice)
```

In this module we will use the Carolina Wren dataset as an example. 
```{r model, warning = F}
Carolina_Wren_Workflow <- workflow(occurrence = CarolinaWrenPO,
                                   covariate = CarolinaWrenRasters,
                                   process = Background(1000),
                                   model = LogisticRegression,
                                   output = PrintMap)
```

We can take a look at our covariate data by calling the accessor function `Process`. This will take the covariate data along with the occurrence data. `str()` will show us the structure of our data, allowing us to confirm the form our data are stored in. 
```{r}
cov <- Process(Carolina_Wren_Workflow)$df
levels(cov) <- c('background', 'presence')
str(cov)
```
In this case, we can see that data type is stored as a factor, and the continuous variables are stored as numeric variables. This is correct. 

Next we might want to check that there is roughly even coverage of our two types of data, or our factor variable. 
```{r}
table(cov$type)
```
There are!

At this point we can begin our data exploration. The first thing to do is to check for outliers in our data. These may indicate data entry errors when using your own data, or access errors,  example. 
Data exploration
A Outliers in Y / Outliers in X
B Collinearity X
C Relationships Y vs X
D Spatial/temporal aspects of sampling design 
E Interactions (is the quality of the data good enough to
                 include them?)
F Zero inflation Y
G Are categorical covariates balanced?

# OUTLIERS?
```{r}
par(mfrow = c(4, 1), mar = c(3, 4, 1, 1))
dotchart(cov$pcCon, main= "Percent Coniferous")
dotchart(cov$pcDec, main= "Percent Deciduous")
dotchart(cov$pcMix, main= "Percent Mixed Forest")
dotchart(cov$pcGr, main= "Percent Grassland")
dev.off()
```

```{r}
# RELATIONSHIPS?
 panel.cor <- function(x, y, digits = 2, cex.cor, ...)
  {
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    # correlation coefficient
    r <- cor(x, y)
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste("r= ", txt, sep = "")
    text(0.5, 0.6, txt)
    
    # p-value calculation
    p <- cor.test(x, y)$p.value
    txt2 <- format(c(p, 0.123456789), digits = digits)[1]
    txt2 <- paste("p= ", txt2, sep = "")
    if(p<0.01) txt2 <- paste("p= ", "<0.01", sep = "")
    text(0.5, 0.4, txt2)
  }

 pairs(cov[,c('pcCon', 'pcDec', 'pcMix', 'pcGr')], upper.panel = panel.cor)
```

```{r}
# Covariate effects
par(mfrow = c(1, 1), mar = c(5, 5, 2, 2), cex.lab = 1.5)
boxplot(cov$pcCon ~ type,
        varwidth = TRUE,
        data = cov,
        xlab     = "Data Type",
        ylab     = "Percent Coniferous")
abline(h = mean(cov$pcCon, na.rm = TRUE),
       lty = 2)
```


